---
title: "Data004: AI Adoption and Digital Capability lvsem Demo"
author: "Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook analyzes the public Mendeley Data dataset by Navara and Prabowo
(2026), **AI Adoption, Digital Capability, and Innovation Capability: The
Mediating Role of Strategic Decision-Making Quality**.

The model follows the dataset description. AI adoption and digital capability
are treated as antecedents of strategic decision-making quality and innovation
capability, while strategic decision-making quality is treated as a mediator.
The analysis uses the local `lvsem` package in the same style as the other
demo folders.

## Packages

```{r packages}
rm(list = ls(all = TRUE))

library(lvsem)
library(readxl)
library(ggplot2)
library(knitr)
```

## Data

```{r load_data}
data_file <- "Excel data respondent.xlsx"
data_candidates <- c(
  file.path("data004", data_file),
  data_file,
  file.path("..", "data004", data_file)
)
data_path <- data_candidates[file.exists(data_candidates)][1]

if (is.na(data_path)) {
  stop("Excel file not found: ", data_file)
}

data_dir <- dirname(data_path)
raw_data <- readxl::read_excel(data_path)
names(raw_data) <- trimws(gsub("[\r\n]+", "\n", names(raw_data)))

dim(raw_data)
head(raw_data)
```

## Model Definition

```{r model_definition}
item_map <- c(
  AIA1 = "I intend to adopt an AI-based system in my business/work because it can improve my business performance and results.",
  AIA2 = "I use AI-based technology to help analyze data or information in my work or business.",
  AIA3 = "The application of AI technology helps improve the efficiency of my work processes or business operations.",
  AIA4 = "I believe that AI systems can provide reliable results or recommendations in my work.",
  AIA5 = "The use of AI technology aligns with the needs and working methods of my business or work.",
  DC1 = "I have good skills in using various digital technologies to support my work/business.",
  DC2 = "I am able to utilize digital technologies (such as software, online platforms, or data systems) to support my work or business activities.\n",
  DC3 = "I have the ability to manage and utilize digital information or data to support work decision-making.",
  DC4 = "I am able to use various digital technologies to improve the efficiency of work processes or business operations.\n",
  DC5 = "I can integrate digital technology with my work processes or business strategies.\n",
  DC6 = "I am able to utilize digital technology to create or develop new ideas, products, or services in my work or business.\n",
  DC7 = "I have the ability to combine digital technology with organizational resources to improve work or business performance.\n",
  DC8 = "I have skills in using digital technology to communicate and collaborate with colleagues or business partners.\n",
  SDQ1 = "I always set clear goals before making strategic decisions in my work or business.",
  SDQ2 = "Before making a strategic decision, I gather information from various relevant sources (both internal and external).\n",
  SDQ3 = "I systematically evaluate several alternative solutions before making a final decision.\n",
  SDQ4 = "In making strategic decisions, I consider the potential risks and the chances of success of each alternative.\n",
  SDQ5 = "My strategic decisions are based on logical and structured analysis, not just intuition.\n",
  IC1 = "Before making strategic decisions, I set measurable and specific decision objectives as a basis for evaluation.\n",
  IC2 = "I evaluate strategic decision alternatives by quantitatively considering the probability of risks and their business consequences.",
  IC3 = "The strategic decisions I make are based on relevant data and analysis, not just personal intuition.\n",
  IC4 = "I make optimal use of the knowledge and information available within the organization before making strategic decisions.\n",
  IC5 = "In the strategic decision-making process, I encourage team participation to improve the quality of alternative evaluations.",
  IC6 = "I re-evaluate previous strategic decisions to improve the quality of subsequent decisions.\n",
  IC7 = "Every strategic decision I make is aligned with the company's long-term strategy and business objectives.\n"
)

item_map <- trimws(gsub("[\r\n]+", "\n", item_map))

missing_items <- setdiff(unname(item_map), names(raw_data))
if (length(missing_items) > 0) {
  stop("Missing expected columns: ", paste(missing_items, collapse = " | "))
}

model_data <- as.data.frame(raw_data[, unname(item_map)])
names(model_data) <- names(item_map)

model_data[] <- lapply(model_data, function(x) {
  x <- as.numeric(as.character(x))
  x[!is.finite(x)] <- NA_real_
  x
})

model_data[] <- lapply(model_data, function(x) {
  ifelse(is.na(x), stats::median(x, na.rm = TRUE), x)
})

blocks <- list(
  AI_Adoption = paste0("AIA", 1:5),
  Digital_Capability = paste0("DC", 1:8),
  Strategic_Decision_Quality = paste0("SDQ", 1:5),
  Innovation_Capability = paste0("IC", 1:7)
)

block_summary <- data.frame(
  block = names(blocks),
  item_count = vapply(blocks, length, integer(1))
)
kable(block_summary, caption = "Measurement blocks")

latent_vars <- names(blocks)
path <- matrix(
  0,
  nrow = length(latent_vars),
  ncol = length(latent_vars),
  dimnames = list(latent_vars, latent_vars)
)

path["Strategic_Decision_Quality", c("AI_Adoption", "Digital_Capability")] <- 1
path["Innovation_Capability", c(
  "AI_Adoption",
  "Digital_Capability",
  "Strategic_Decision_Quality"
)] <- 1

path

score_relation <- path + t(path)
score_relation
```

The final seven manifest variables are labeled as `Innovation_Capability` to
match the dataset title and description. Their literal wording is strongly
decision-quality oriented, so the construct interpretation should be checked
against the authors' intended instrument.

## Export Clean Data

```{r export_clean_data}
item_dictionary <- data.frame(
  construct = rep(names(blocks), vapply(blocks, length, integer(1))),
  item = unlist(blocks, use.names = FALSE),
  source_column = unname(item_map[unlist(blocks, use.names = FALSE)]),
  row.names = NULL
)

write.csv(model_data, file.path(data_dir, "data004_ai_adoption_clean.csv"), row.names = FALSE)
write.csv(item_dictionary, file.path(data_dir, "Data_Dictionary.csv"), row.names = FALSE)

data.frame(
  samples = nrow(model_data),
  variables = ncol(model_data)
)
```

## Latent Variable Scores

```{r scores}
lv_scores <- make_lv_scores(model_data, blocks, score_relation)

head(lv_scores)
round(cor(lv_scores), 3)

write.csv(lv_scores, file.path(data_dir, "data004_latent_scores.csv"), row.names = FALSE)
```

## Structural Model

```{r structural_model}
path_coefficients <- estimate_paths(lv_scores, path, scale_scores = FALSE)
path_effects <- effects_from_beta(path_coefficients)

path_coefficients
path_effects

endogenous_vars <- rownames(path)[rowSums(path) > 0]
r2_values <- sapply(endogenous_vars, function(dv) {
  preds <- names(which(path[dv, ] == 1))
  fit <- lm(
    as.formula(paste(dv, "~", paste(preds, collapse = " + "))),
    data = as.data.frame(lv_scores)
  )
  summary(fit)$r.squared
})

r2_table <- data.frame(
  construct = names(r2_values),
  R2 = as.numeric(r2_values),
  row.names = NULL
)

kable(
  data.frame(
    construct = r2_table$construct,
    R2 = round(r2_table$R2, 3)
  ),
  caption = "R2 values"
)

write.csv(r2_table, file.path(data_dir, "data004_r2.csv"), row.names = FALSE)
```

## Bootstrap

```{r bootstrap}
set.seed(123)
boot_paths <- bootstrap_paths(
  model_data,
  blocks,
  path,
  br = 1000,
  seed = 123,
  score_matrix = score_relation
)

kable(
  data.frame(
    path = boot_paths$path,
    round(boot_paths[, c("beta", "boot_se", "t_value")], 3)
  ),
  caption = "Bootstrap path estimates"
)

write.csv(
  boot_paths,
  file.path(data_dir, "data004_path_coefficients_bootstrap.csv"),
  row.names = FALSE
)
```

## Measurement Model

```{r measurement}
measures <- measurement_summary(model_data, blocks, lv_scores)

kable(
  data.frame(
    block = measures$reliability$block,
    round(measures$reliability[, c("alpha", "composite_reliability", "ave")], 3)
  ),
  caption = "Reliability and AVE"
)

kable(
  data.frame(
    block = measures$loadings$block,
    item = measures$loadings$item,
    loading = round(measures$loadings$loading, 3)
  ),
  caption = "Outer loadings"
)

write.csv(
  measures$reliability,
  file.path(data_dir, "data004_reliability_ave.csv"),
  row.names = FALSE
)
write.csv(
  measures$loadings,
  file.path(data_dir, "data004_outer_loadings.csv"),
  row.names = FALSE
)
```

## Model Plot

```{r model_plot, fig.width=12, fig.height=7}
node_df <- data.frame(
  block = c(
    "AI_Adoption",
    "Digital_Capability",
    "Strategic_Decision_Quality",
    "Innovation_Capability"
  ),
  label = c(
    "AI Adoption\n(AI_Adoption)",
    "Digital Capability\n(Digital_Capability)",
    "Strategic Decision\nQuality\n(Strategic_Decision_Quality)",
    "Innovation\nCapability\n(Innovation_Capability)"
  ),
  x = c(0, 0, 3.0, 6.2),
  y = c(2.2, 0.4, 1.3, 1.3),
  stringsAsFactors = FALSE
)

edge_df <- boot_paths
edge_split <- strsplit(edge_df$path, "_to_")
edge_df$from <- vapply(edge_split, `[`, character(1), 1)
edge_df$to <- vapply(edge_split, `[`, character(1), 2)
edge_df$label <- sprintf("beta = %.3f\nT = %.3f", edge_df$beta, edge_df$t_value)

edge_df <- merge(
  edge_df,
  node_df[, c("block", "x", "y")],
  by.x = "from",
  by.y = "block"
)
names(edge_df)[names(edge_df) %in% c("x", "y")] <- c("x_from", "y_from")

edge_df <- merge(
  edge_df,
  node_df[, c("block", "x", "y")],
  by.x = "to",
  by.y = "block"
)
names(edge_df)[names(edge_df) %in% c("x", "y")] <- c("x_to", "y_to")

edge_df$x_label <- c(1.35, 1.35, 3.25, 3.25, 4.70)[match(
  edge_df$path,
  c(
    "AI_Adoption_to_Strategic_Decision_Quality",
    "Digital_Capability_to_Strategic_Decision_Quality",
    "AI_Adoption_to_Innovation_Capability",
    "Digital_Capability_to_Innovation_Capability",
    "Strategic_Decision_Quality_to_Innovation_Capability"
  )
)]
edge_df$y_label <- c(1.85, 0.72, 2.62, 0.18, 1.72)[match(
  edge_df$path,
  c(
    "AI_Adoption_to_Strategic_Decision_Quality",
    "Digital_Capability_to_Strategic_Decision_Quality",
    "AI_Adoption_to_Innovation_Capability",
    "Digital_Capability_to_Innovation_Capability",
    "Strategic_Decision_Quality_to_Innovation_Capability"
  )
)]

r2_df <- data.frame(
  block = r2_table$construct,
  r2_label = sprintf("R2 = %.3f", r2_table$R2),
  stringsAsFactors = FALSE
)
r2_df <- merge(r2_df, node_df[, c("block", "x", "y")], by = "block")
r2_df$y_label <- ifelse(
  r2_df$block == "Strategic_Decision_Quality",
  r2_df$y + 0.62,
  r2_df$y - 0.55
)

model_plot <- ggplot() +
  geom_segment(
    data = edge_df,
    aes(
      x = x_from + 0.58,
      y = y_from,
      xend = x_to - 0.58,
      yend = y_to
    ),
    arrow = arrow(length = unit(0.20, "cm")),
    linewidth = 0.75,
    color = "#374151",
    alpha = 0.90
  ) +
  geom_label(
    data = node_df,
    aes(x = x, y = y, label = label),
    size = 3.8,
    fontface = "bold",
    label.size = 0.35,
    fill = "#ffffff",
    color = "#111827"
  ) +
  geom_label(
    data = edge_df,
    aes(x = x_label, y = y_label, label = label),
    size = 2.9,
    label.size = 0.20,
    fill = "#ffffff",
    color = "#111827"
  ) +
  geom_label(
    data = r2_df,
    aes(x = x, y = y_label, label = r2_label),
    size = 3.0,
    label.size = 0.25,
    fill = "#eef7f2",
    color = "#111827"
  ) +
  coord_cartesian(xlim = c(-0.9, 7.1), ylim = c(-0.5, 3.0), clip = "off") +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "#ffffff", color = NA),
    panel.background = element_rect(fill = "#ffffff", color = NA),
    plot.title = element_text(color = "#111827", face = "bold", hjust = 0.5)
  ) +
  labs(title = "AI Adoption and Digital Capability lvsem Model")

model_plot

ggsave(
  file.path(data_dir, "data004_lvsem_ai_adoption_model.png"),
  model_plot,
  width = 12,
  height = 7,
  dpi = 180
)
```

## Final Output

```{r final_output}
cat("Sample size:", nrow(model_data), "\n")
print(round(r2_values, 3))
```

## Reference

- Navara, D., & Prabowo, P. (2026). *AI Adoption, Digital Capability, and
  Innovation Capability: The Mediating Role of Strategic Decision-Making
  Quality*. Mendeley Data, V2. https://doi.org/10.17632/fs78cfs84v.2
- Local package: `lvsem`
- Source file: `Excel data respondent.xlsx`
